---
title: Introducing parmsurvfit Package - Simple Parametric Survival Analysis with R
author:
  - name: Victor Wilson
    affiliation: California Polytechnic State University, San Luis Obispo - Statistics Department
    email:  victorjw26@yahoo.com
  - name: Ashley Jacobson
    affiliation: California Polytechnic State University, San Luis Obispo - Statistics Department
    email:  ashleypjacobson@gmail.com
  - name: Shannon Pileggi
    affiliation: California Polytechnic State University, San Luis Obispo - Statistics Department
    email:  spileggi@calpoly.edu
abstract: >
  This article introduces the R package parmsurvfit, which executes basic parametric survival analysis
  techniques similar to those in Minitab. Among these are plotting hazard, cumulative hazard, survival, and  density curves, computing survival probabilites, and computing summary statistics based on a specified parametric distribution. We describe appropriate usage of these functions, interpretation of output, and provide examples of how to utilize these functions in real-world datasets. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

## Introduction

<<<<<<< HEAD
Survival analysis is a branch of statistics that primarily deals with analyzing the time until an event of interest occurs. This event could be a variety of different things such as death, development of disease, or first score of a soccer game. Observations in survival analysis may also be described as censored, which occurs when an observation's survival time is incomplete. The most common way that this occurs is through right censoring, which occurs when a subject does not experience the event of interest within the duration of the study. Right censoring can also occur if a subject drops out before the end of the study and does not experience the event of interest. Due to the inherent issue of censoring that is typically found in datasets involving survival analysis, computations and analyses can be difficult to carry out with many standard functions available in R, as the majority of these do not account for censored data. The censored data collected is of value and we cannot merely eliminate the observations which have censored data. 

Some of the most popular techniques and statistics utilized when carrying out a survival analysis are computing what are known as the survival and hazard functions \citep{Kleinbaum2012}. The survival function is important because it gives the probability of surviving (also known as not experiencing the event of interest) beyond any given time $t$.  Similarly, the hazard function is also useful to compute because it gives the conditional probability that the subject will experience the event in the next instance of time, given that they have survived up until the specified point in time. Statistics that are commonly reported include median survival time, mean survival time, and percentiles of survival time. In this package, all of the functions that we developed utilize parametric methods of survival analysis, which assumes that the distribution of the survival times follows a known probability distribution. 
=======
Survival analysis is a branch of statistics that primarily deals with analyzing the time until   an event of interest occurs. This event could be a variety of different things such as death,   development of disease, or first score of a soccer game. Observations in survival analysis may   also be described as censored, which occurs when an observation's survival time is incomplete.  The most common way that this occurs is through right censoring, which occurs when a subject   does not experience the event of interest within the duration of the study. Right censoring can also occur if a subject drops out before the end of the study and does not experience the event of interest. Due to the inherent issue of censoring that is typically found in datasets involving survival analysis, computations and analyses can be difficult to carry out with many standard functions available in R, as the majority of these do not account for censored data. The censored data collected is of value and we cannot merely eliminate the observations which have censored data. 

<<<<<<< HEAD
Some of the most popular techniques and statistics utilized when carrying out a survival analysis are computing what are known as the survival and hazard functions. The survival function is important because it gives the probability of surviving (also known as not experiencing the event of interest) beyond any given time $t$.  Similarly, the hazard function is also useful to compute because it gives the conditional probability that the subject will experience the event in the next instance of time, given that they have survived up until the specified point in time. Other popular statistics that are utilized are median survival time, mean survival time, and percentiles of survival time. In this package, all of the functions that we developed utilize parametric methods of survival analysis, which assumes that the distribution of the survival times follows a known probability distribution. 
=======
Some of the most popular techniques and statistics utilized when carrying out a survival analysis are computing what are known as the survival and hazard functions \citep{Kleinbaum2012}. The survival function is important because it gives the probability of surviving (also known as not experiencing the event of interest) beyond any given time $t$.  Similarly, the hazard function is also useful to compute because it gives the conditional probability that the subject will experience the event in the next instance of time, given that they have survived up until the specified point in time. Statistics that are commonly reported include median survival time, mean survival time, and percentiles of survival time. In this package, all of the functions that we developed utilize parametric methods of survival analysis, which assumes that the distribution of the survival times follows a known probability distribution. 
>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d
>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

Currently, R does have many survival packages that address non-parametric survival analysis, such as the \pkg{survival} package.   Moreover, R does have some packages that aid in estimation for parametric survival analysis, including \pkg{fitdistrplus}. However, Minitab has very concise and easy to utilize functions for computing and displaying many parametric survival statistics and plots, but this same output is not readily available in any single one package in R, or in some cases not available at all. Thus, we decided to develop a package that emulates the output found in Minitab for parametric survival analysis, which contains all of these commonly utilized statistics and plots.

This paper describes the functions that the \pkg{parmsurvfit} package contains, how the data is formatted in order to utilize these functions, and what the output of these functions represent. There are three major groups of functions that we created: assessing fit, survival functions (density, hazard, cumulative hazard, and survival), and computing statistics (mean, median, survival probabilities). The majority of this paper will be organized following these groups of functions. 


<<<<<<< HEAD
The \code{fit_data} function produces maximum likelihood estimates for right censored data based on the input distribution.  The \code{fit_data} function utilizes the \code{fitdistcens} function in the [fitdistrplus](https://cran.r-project.org/web/packages/fitdistrplus/index.html) package, but allows for more intuitive input of right-censored data than as specified with \code{fitdistcens}. The \code{fit\_data} function in the \pkg{parmsurvfit} package inputs two variables: \code{time} and \code{censor}. The \code{time} variable contains the time-to-event variable, while the \code{censor} variable indicates whether right censoring is present (0 corresponds to censored data and 1 corresponds to complete data). Furthermore, the user specifies the desired parametric distribution in \code{dist} by inputting the base name of the distribution as a character string.  For example, to utilize the normal distribution you would specify "norm" as it is the base of \code{dnorm}, \code{pnorm}, etc.  Commonly utilized distributions for survival analysis include Weibull ("weibull"), log-normal ("lnorm"), exponential ("exp"), and logistic ("logis") distributions.  The function also takes in an optional grouping variable, which fits the data for each group individually. The function returns an object of class "fitdistcens", and if there's a grouping variable it returns a list of objects of class "fitdistcens".
=======
<<<<<<< HEAD
## Assessing fit
Since all of the functions available in this package assume that the survival data follows a  
known parametric distribution, it is important to have a method to analyze how well our assumed   model fits the data. Utilizing such methods will allow us choose a distribution that adequately   fits the data. Some common methods used to assess goodness of fit are viewing a histogram of the  data, Q-Q (Quantile-Quantile) plots, the Anderson-Darling Test.   

### Fitting right censored survival data

As mentioned previously, this function is very similar to the function \code{fitdistcens} found in the [fitdistrplus](https://cran.r-project.org/web/packages/fitdistrplus/index.html) package, which computes the Maximum Likelihood Estimates (MLEs) for right-censored data. The \code{fitdistcens} function requires data to be organized into two columns, right and left. The right column indicates a start time, and the left column indicates an end time. For example, if a time is right censored, then the left column would contain the time, and the right column would contain NA. This way of organizing survival data allows for different types of censoring to be unambiguous, but since the \pkg{parmsurvfit} package will only handle right censored data, the \code{fit\_data} function makes it so that the user doesn't have to reorganize data. 
>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

This function takes in two required columns, a Time column and a Censor column. The time column contains the time-to-event variable, while the censor column indicates whether right censoring is present (0 corresponds to censored data and 1 corresponds to complete data). The function also takes in an optional grouping variable, which fits the data for each group individually. The function returns an object of class \samp{fitdistcens}, and if there's a grouping variable it returns a list of objects of class \samp{fitdistcens}. Several of the other functions available in this package also utilize arguments that are similar to the ones found in this function. 
=======
Then the \code{fit\_data} function produces maximum likelihood estimates for right censored data based on the input distribution.  The \code{fit\_data} function utilizes the \code{fitdistcens} function in the [fitdistrplus](https://cran.r-project.org/web/packages/fitdistrplus/index.html) package, but allows for more intuitive input of right-censored data than as specified with \code{fitdistcens}. The \code{fit\_data} function in the \pkg{parmsurvfit} package inputs two variables: \code{time} and \code{censor}. The \code{time} variable contains the time-to-event variable, while the \code{censor} variable indicates whether right censoring is present (0 corresponds to censored data and 1 corresponds to complete data). Furthermore, the user specifies the desired parametric distribution in \code{dist} by inputting the base name of the distribution as a character string.  For example, to utilize the normal distribution you would specify "norm" as it is the base of \code{dnorm}, \code{pnorm}, etc.  Commonly utilized distributions for survival analysis include Weibull ("weibull"), log-normal ("lnorm"), exponential ("exp"), and logistic ("logis") distributions.  The function also takes in an optional grouping variable, which fits the data for each group individually. The function returns an object of class "fitdistcens", and if there's a grouping variable it returns a list of objects of class "fitdistcens".

>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d

### Example

In this example, we fit the Weibull distribution to the \file{firstdrink} data set where the time to event variable is age and the variable that indicates censoring status is censor.  The maximum likelihood estimates of the location and scale parameters are returned.

```{r, echo=FALSE}
library(parmsurvfit)
datapath<-"/Users/Victor/Survival Packages Frost Research/Data sets/"
Firstdrink<-read.table(paste0(datapath,"Firstdrink.txt"),header=T)
fit_data(data = Firstdrink, dist = "weibull", time = "Age", censor = "Censor")
```

<<<<<<< HEAD
## Assessing fit

Since all of the functions available in this package assume that the survival data follows a known parametric distribution, it is important to have a method to analyze how well our assumed model fits the data. Utilizing such methods will allow us choose a distribution that adequately fits the data. A few common methods used to assess goodness of fit are viewing histograms and density plots, viewing Q-Q (Quantile-Quantile) plots and the Anderson-Darling test statistic.

### Density plots/histograms

Histograms and density plots are another useful way to assess the fit of your data. Being able to view your data in a histogram, and then overlay the function you are using to model your data is a great way to visualize how accurate the parameterization is for your data. 

The \code{plot\_density} function creates a histogram of the data and overlays the density function of a fitted parametric distribution. Parameters estimates for the specified parametric distribution are provided as well. This function also supports the ability to plot separate histograms and density functions for each level of a grouping variable. Below, we fit the weibull distribution to age until first drink by each value for gender using the \file{firstdrink} data set.  Three plots are produced, each based on separate MLE estimates: a plot for males (level = 1), females (level = 2), and overall.  In these plots, all time to event data are plotted regardless of censoring status.

```{r}
plot_density(data = firstdrink, dist = "weibull", time = "age", censor = "censor", by = "gender")
```



### Q-Q plots

In Q-Q plots the hypothesized quantiles are plotted on the $y$-axis and the observed quantiles are plotted on the $x$-axis. A $y=x$ line is typically included in these plots, because if the observed data fit the hypothesized distribution perfectly, all of the points would lie exactly on this diagonal line. The \code{plot\_qqsurv} function creates a quantile-quantile plot of right-censored data given that it follows a specified distribution. 

**May need to modify: The points are plotted according to the median rank method \citep{Minitabqq}.**

```{r}
plot_qqsurv(data = firstdrink, dist = "weibull", time = "age", censor = "censor")
```
The plot above displays a Q-Q plot for the the \file{firstdrink} data set to see how well the Weibull distribution fits the data. As seen in the plot above, there are some deviations from the provided $y=x$ line, indicates that an Weibull distribution may not be an ideal fit for the data.   

### Anderson-Darling test statistic

While creating Q-Q plots are a great way to visualize how a particular distribution may fit the data, it can be difficult at times to definitively decide how well the plot fits the data. The Anderson-Darling test statistic provides a numerical test statistic that measures how well the data fits a particular distribution, such that lower values indicate a better fit. Computation of the test statistic adhered to Minitab's documentation, utilizing the median rank plotting method \citep{Minitabgof}.
=======
### Density plots/histograms
The \code{plot\_density} function creates a histogram of the data and overlays the density function of a fitted parametric distribution. Parameter estimates for the specified parametric distribution are provided as well. This function also supports the ability to plot separate histograms and density functions for each level of a grouping variable. An example of this function is shown below:
>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

```{r}
plot_density(Firstdrink, "norm", time = "Age", by = "Gender")
```
We ran the \code{plot\_density} function, utilizing the \file{Firstdrink.txt} dataset available in our package. This dataset contains data on the age of first consumption of an alcoholic beverage for 1000 individuals. As seen above, a separate histogram and density plot was created for males and females.

### Q-Q plots
As mentioned before, Q-Q plots are a very popular method used to evaluate the fit between two probability distributions. In these plots, the hypothesized quantiles are plotted on one axis and the observed quantiles are plotted on the other axis. A *y=x* line is typically included in these plots, because if the observed data fit the hypothesized distribution perfectly, all of the points would lie exactly on this line. We developed the \code{plot\_qqsurv} function to create a quantile-quantile plot of right-censored data given that it follows a specified distribution. 

<<<<<<< HEAD
## Survival Functions

Several functions are utilized in survival analysis to compute statistics about the event of interest. This section provides an overview of the survival functions available in this package. Some of the most common functions used in survival analysis are the survival function, the hazard function, and the cumulative hazard function. We designed these functions with an intent to have the output displayed be very easy to read and interpret. Table \ref{table:functions} lists each function and its relationship to other functions, as well as the formula used to compute each function. 
=======
```{r}
plot_qqsurv(Firstdrink, "exp", time="Age")
```
We can create a Q-Q plot for the the Firstdrink data set to see how well an Exponential distribution fits the data. As seen in the plot above, there are some deviations from the provided *y=x* line, indicates that an Exponential distribution may not be an ideal fit for the data. 

### Anderson-Darling Test
While creating Q-Q plots are a great way to visualize how a particular distribution may fit the data, it can be difficult at times to definitively decide how well the plot fits the data. The Anderson-Darling test provides a numerical test statistic that measures how well the data fits a particular distribution. **CITE MINITAB HELP PAGE** 



## Survival Functions
This section introduces an overview of the many types of survival functions that are able to be displayed via this package. Some of the most common functions used in survival analysis are the survival function, the hazard function, and the cumulative hazard function. 

<<<<<<< HEAD
We designed functions within this package to display a plot for each of the aforementioned survival functions with an intent to have the output displayed be very easy to read and interpret. Below is a list of each function and it's relationship to other functions, as well as the formula used to compute each function. 
=======
## Plots for survival analysis
>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d

This section provides overview of the plots available in this package. Some of the most common plots used in survival analysis are density plots, survival plots, hazard plots, and cumulative hazard plots. We designed these functions with an intent to have the output displayed be very easy to read and interpret. Table \ref{table:functions} lists each function and its relationship to other functions, as well as the formula used to compute each function. 
>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

\begin{table}
\begin{tabular}{ll}
\hline
Function & Relationships  \\
\hline
PDF & ${f(t)=\frac{d}{dt}F(t)}$\\
CDF  & ${F(t)=\int_0^t f(y)dy}$\\
Survival & ${S(t)=1-F(t)=\exp[-H(t)]=\exp[-\int_0^th(y)dy]}$ \\
Hazard & ${h(t)=f(t)/S(t)=-\frac{d}{dt}\ln[S(t)]}$ \\
Cum. Haz. & ${H(t)=\int_0^t h(y)dy=-\ln[S(t)]}$\\
\hline
\end{tabular}
\label{table:functions}
\end{table}

<<<<<<< HEAD
=======
<<<<<<< HEAD

### Survival plots
Survival plots are used to estimate the proportion of subjects that survive beyond a specified time **t**. We were motivated to create the function \code{plot\_surv} in an attempt to create hazard plots that are easy to produce, when dealing with data set up for survival analysis. This function plots the survival curve of right censored data given that it follows a specified parametric distribution. Some examples of the distributions that this function supports are the Weibull, Log-Normal, Exponential, Normal, and Logistic distributions. This function also provides the option to plot by a grouping variable, which if specified, displays separate curves for each group of the specified variable. In these plots, survival time is plotted on the x-axis, while survival probability is plotted on the y-axis.
=======
### Density plots/histograms

The \code{plot\_density} function creates a histogram of the data and overlays the density function of a fitted parametric distribution. Parameters estimates for the specified parametric distribution are provided as well. This function also supports the ability to plot separate histograms and density functions for each level of a grouping variable. Below, we fit the weibull distribution to age until first drink by each value for gender using the \file{firstdrink} data set.  Three plots are produced, each based on separate MLE estimates: a plot for males (level = 1), females (level = 2), and overall.  In these plots, all time to event data are plotted regardless of censoring status.

```{r}
plot_density(data = Firstdrink, dist = "weibull", time = "Age", censor = "Censor", by = "Gender")
```
>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

### Survival function

The survival function $S(t)$ estimates the proportion of subjects that survive beyond a specified time $t$. The \code{plot\_surv} function  plots the survival curve of right censored data given a specified parametric distribution.  This function also provides the option to plot by a grouping variable, which if specified, displays separate curves for each value of the specified grouping variable. In these plots, survival time is plotted on the $x$-axis, while survival probability is plotted on the $y$-axis.
>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d




```{r, echo=FALSE}
plot_surv(data = Firstdrink, dist = "weibull", time = "Age", censor = "Censor", by = "Gender")
```
<<<<<<< HEAD
In this example, we fit a Weibull distribution to the 'Firstdrink' dataset, grouping by the Gender variable once again. As seen in the plot above, two different survival curves were plotted. The blue line represents the estimated survival curve for males, while the red line represents the estimated survival curve for females. From this plot, we see that the survival curve for females is consistently above the survival curve for males throughout all points in time. Due to this, we can conclude that males tend to experience their first drink of alcohol before males do. 


=======

### Hazard function

Hazard plots, denoted $h(t)$, are used to display the conditional risk that a subject will experience the event of interest in the next instant of time, given that the subject has survived beyond a certain time $t$. For example, if we observe that a person has survived for 17 years without first trying alcohol, the hazard function would estimate the risk that the person will experience their first drink of alcohol in the next short instant of time, based on the fact that the person has already survived 17 years alcohol free. Hazard is not probability, and therefore can take values greater than one.  Moreover, the hazard function can be both increasing or decreasing.  The \code{plot\_haz} function plots the hazard function based on a specified parametric distribution, with the option to include a grouping variable. 
>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d


```{r, echo=FALSE}
plot_haz(data = Firstdrink, dist = "weibull", time = "Age", censor = "Censor",  by ="Gender")
```
<<<<<<< HEAD
From this plot above, also using the 'Firstdrink' dataset, we can see that as males continue to survive, their risk of experiencing the event of interest in the next instant of time dramatically increases. Similarly, females also seem to have a greater risk of experiencing the event of interest as they survive longer, but their risk is lower than that of males. 

### Cumulative hazard function

While hazard plots are usually useful in assessing a subject's risk of experiencing the event of interest in the next moment of time, these plots can be difficult to read and understand at times. Sometimes, the changes in hazard are very subtle, making it difficult to describe periods of increasing and decreasing risk. In order to accurately assess how hazard rates change over time, we investigate the accumulation of hazard rates over time, known as cumulative hazard. The cumulative hazard function, denoted $H(t)$, is the total accumulated risk of experiencing an event up to time $t$.  

Since the cumulative hazard function is an accumulation of rates, it is important to note that this function is non-decreasing and is hardly ever remains constant by nature. The \code{plot\_cumhaz} function displays cumulative hazard plots, given that the data follows a specified parametric distribution. The functionality of this function is nearly identical to that of \code{plot\_haz}, with the only distinction being that it plots cumulative hazard curves instead of hazard curves. 
>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d


```{r, echo=FALSE}
plot_cumhaz(data = Firstdrink, dist = "weibull", time = "Age", censor = "Censor",  by = "Gender")
```
As expected, the cumulative hazard function is increasing for both males and females. Here, the total accumulated risk of experiencing the first drink of alcohol is greater for males compared to females 

<<<<<<< HEAD
## Statistics/Computations
=======
<<<<<<< HEAD
## Statistics/Computations
While viewing plots such as those explained above are very useful in survival analysis, they only tell half of the story. In order to carry out a complete survival analysis, we must also compute statistics in order to supplement our plots. Some of the most common statistics utilized in parametric survival analysis are survival probabilities and typical summary statistics such as the mean, median, standard deviation, and percentiles of survival time. 
>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b


### Computing survival probabilites

<<<<<<< HEAD
A survival probably estimates the probability that a subject will not have experienced the event of interest beyond a specified time $t$. We developed the function \code{surv\_prob} to compute probability of survival beyond time $t$, given that the data follows a specified parametric distribution. 
=======
A survival probability estimates the probability that a subject will not have experienced the event of interest beyond a specified time $t$. We developed the function \code{surv\_prob} to compute probability of survival beyond time $t$, given that the data follows a specified parametric distribution. 
>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d
>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

```{r}
surv_prob(data = Firstdrink, dist = "weibull", num = 30, time = "Age", censor = "Censor")
```
<<<<<<< HEAD
Utilizing the firstdrink data set and fitting a weibull distribution to the data, the estimated probability that a person survives for 30 years without having their first drink of alcohol is roughly 5%. 
=======
<<<<<<< HEAD
As seen in the output from the function above, utilizing the 'Firstdrink' data set and fitting a log-normal parametric distribution to the data, the estimated probability that a person survives for 30 years without having their first drink of alcohol is roughly 5%. 

>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

### Computing summary statistics

The \code{surv_summary} function estimates various summary statistics, including mean, median, standard deviation, and percentiles of survival time given that the data follows a specified parametric distribution. This function also supports the option to provide separate summary statistics for each level of a grouping variable, if desired.  
<<<<<<< HEAD
=======
>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d
>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

```{r}
surv_summary(data = Firstdrink, dist = "weibull", time = "Age", censor = "Censor", by = "Gender")
```
<<<<<<< HEAD
As seen above, after specifying the grouping variable of gender, two separate tables were produced, one for males and one for females. We can see that the mean survival time for males (16.2 years) is less than the mean survival time for females (18.5 years). 

=======
<<<<<<< HEAD

=======
As seen above, after specifying the grouping variable of gender, two separate tables were produced, one for males and one for females. We can see that the mean survival time for males (16.2 years) is less than the mean survival time for females (18.5 years). 



## Conclusion
<<<<<<< HEAD
The R package \pkg{parmsurvfit} allows for parametric survival analysis methods involving right-censored data to be easily computed in R. The overall goal of developing this package was to provide a central package for R users to find typical methods used in Survival Analysis such as computing survival probabilities, create survival and hazard plots, and assess goodness of fit of a parametric distribution fit to a dataset. 

>>>>>>> 79f9d4625dfaa583ef8dc3c6d291bf0216190e2b

\bibliography{Simple Survival Analysis}
=======

The R package \pkg{parmsurvfit} allows for simple parametric survival analysis methods involving right-censored data to be easily computed in R. The overall goal of developing this package was to provide a central package for R users to utilize typical parametric survival analysis methods such as computing survival probabilities, creating survival and hazard plots, and assessing goodness of fit of a parametric distribution fit to a dataset with right-censored observations. 

## Acknowledgments 

This package development was funded by the Bill and Linda Frost fund at California Polytechnic State University, San Luis Obispo. We would also like to thank Jeff Sklar for his input and guidance.


\bibliography{RJreferences}
>>>>>>> fab9f94e6d9225cfa4ed2576d8985a391a7b604d
