---
title: Introducing parmsurvfit Package - Simple Parametric Survival Analysis with R
author:
  - name: Victor Wilson
    affiliation: California Polytechnic State University, San Luis Obispo - Statistics Department
    email: victorjw26@yahoo.com
  
  - name: Ashley Jacobson
    affiliation: California Polytechnic State University, San Luis Obispo - Statistics Department
    email:  ashleypjacobson@gmail.com
  
  - name: Shannon Pileggi
    affiliation: California Polytechnic State University, San Luis Obispo - Statistics Department
    email:  spileggi@calpoly.edu

abstract: >
  This article introduces the R package \CRANpkg{parmsurvfit}, which executes basic parametric survival analysis techniques similar to those in Minitab. Among these are plotting hazard, cumulative hazard, survival, and  density curves, computing survival probabilites, and computing summary statistics based on a specified parametric distribution. We describe appropriate usage of these functions, interpretation of output, and provide examples of how to utilize these functions in real-world datasets. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

# Introduction

Survival analysis is a branch of statistics that primarily deals with analyzing the
time until an event of interest occurs. This event could be a variety of different
endpoints such as death, development of disease, or first score of a soccer game.
Observations in survival analysis may also be described as censored, which occurs when
an observation's actual time to event is unknown, but is known to be within some
specified range. For example, if a subject does not experience the event of interest
prior to study completion, then their "observed" event time (time until study
completion) is less than their "actual" event time.  This particular case is known as
right-censoring, which is the most common form. Right censoring can also occur if a
subject drops out before the end of the study and does not experience the event of
interest. 

Typically, survival analysis begins with computing what are known as the survival and 
hazard functions \citep{Kleinbaum2012}. The survival function estimates the 
probability of surviving (also known as not experiencing the event of interest) beyond
any giventime $t$.  The hazard function computes the conditional probability that the 
subject will experience the event in the next instance of time, given that they have 
survived up until the specified point in time. Summary statistics that are commonly 
reported include median survival time, mean survival time, and percentiles of survival
time. These functions and statistics can be computed by either parametric or 
nonparametric techniques.  In the \pkg{parmsurvfit} package, all of the functions that
we developed utilize parametric methods of survival analysis, which assumes that the
distribution of the survival times follows a known probability distribution.

R has many survival packages that address non-parametric survival analysis, such as 
the \CRANpkg{survival} package.  However, prior to the \pkg{parmsurvfit} package, 
parametric estimation of survival functions and statistics was not easily attained. 
The \pkg{parmsurvfit} package attempts to emulate the ease and functionality of 
parametric survival analysis features available in Minitab.   

This paper describes example survival data available in the \pkg{parmsurvfit} package,
the functions that the package contains, how the data is formatted in order to utilize
these functions, and what the output of these functions represent. There are three
major groups of functions that we created: assessing fit, survival functions (density,
survival, hazard, cumulative hazard), and computing statistics (mean, median, survival
probabilities). Explanations are presented according to these groups of functions. 

# Data

The \pkg{parmsurvfit} package contains five data sets with observations subject to
right-censoring (\code{aggressive}, \code{firstdrink}, \code{graduate}, \code{oscars},
and \code{rearrest}).  Subsequent examples in this paper are based on the
\code{firstdrink} data set, which contains 1,000 records from the 1990-1992 National
Comorbidity Survey regarding age at first drink of alcohol (\code{age}).  An
observation is recorded as complete (\code{censor = 1}) if the age at first drink of
alcohol is known.  An observation is recorded as incomplete, or right-censored
(\code{censor = 0}) if the subject had not yet consumed an alcoholic beverage at the
time of the study execution.  In this case, the subject's "actual" event time (age at
first drink) is only known to exceed their "observed" event time (age at time of
study). This data set also includes a \code{gender} variable such that \code{1}
corresponds to males and \code{2} corresponds to females.


# Fitting right-censored survival data


The \code{fit\_data} function produces maximum likelihood estimates (MLE) for right 
censored data based on the input distribution.  The \code{fit\_data} function utilizes
the \code{fitdistcens} function in the \CRANpkg{fitdistrplus} package, but allows for 
more intuitive input of right-censored data than as specified with \code{fitdistcens} 
(which allows input of other types of censoring). The \code{fit\_data} function in the
\pkg{parmsurvfit} package inputs two variables: \code{time} and \code{censor}. The 
\code{time} variable contains the time-to-event variable, while the \code{censor} 
variable indicates whether right censoring is present (0 corresponds to censored data 
and 1 corresponds to complete data). Furthermore, the user specifies the desired 
parametric distribution in \code{dist} by inputting the base name of the distribution 
as a character string.  For example, to utilize the normal distribution you would 
specify \samp{norm} as it is the base of \code{dnorm}, \code{pnorm}, etc.  Commonly 
utilized distributions for survival analysis include Weibull (\samp{weibull}), 
log-normal (\samp{lnorm}), exponential (\samp{exp}), and logistic (\samp{logis}) 
distributions.  The function also takes in an optional grouping variable, which fits 
the data for each group individually. The function returns an object of class 
\samp{fitdistcens}, and if there is a grouping variable it returns a list of objects 
of class \samp{fitdistcens}.  

## Example

In this example, we fit the Weibull distribution to the \file{firstdrink} data set 
where the time to event variable is \code{age} and the variable that indicates 
censoring status is \code{censor}.  The maximum likelihood estimates of the location 
and scale parameters are returned.

```{r fit}
library(parmsurvfit)
fit_data(data = firstdrink, dist = "weibull", time = "age", censor = "censor")
```

# Assessing fit

Since all of the functions available in this package assume that the survival data 
follows a known parametric distribution, we presents methods to evaluate how well the 
assumed model fits the data. Some common methods used to assess goodness of fit are 
viewing a histogram of the data, Q-Q (Quantile-Quantile) plots, the Anderson-Darling 
test statistic. 

## Histograms with density curves

The \code{plot\_density} function produces histograms with overlayed density curves to
allow a user to visually assess fit of a parametric distribution to data.  Parameter 
estimates for the specified parametric distribution are provided as well. This 
function also supports the ability to plot separate histograms and density functions 
for each level of a grouping variable. Below, we fit the weibull distribution to age 
until first drink by gender.  Three plots are produced, each based on their respective 
MLE's: a plot for males (\code{level = 1}), females (\code{level = 2}), and overall.  
In these plots, all time to event data are plotted regardless of censoring status.


```{r densityplot}
plot_density(data = firstdrink, dist = "weibull", time = "age", censor = "censor", by = "gender")
```

Here, we observe that the Weibull distribution could possibly be too flat for the 
distribution of age until first drink of alcohol.


## Q-Q plots

**Quantiles vs percentileS!!!**

The \code{plot\_qqsurv} function creates a quantile-quantile plot of right-censored 
data given that it follows a specified distribution. In typical Q-Q plots the 
hypothesized (theoretical) quantiles are plotted on the $y$-axis and the observed 
(sample) quantiles are plotted on the $x$-axis. A $y=x$ line is included in these 
plots, because if the observed data fit the hypothesized distribution perfectly, all 
of the points would lie exactly on this diagonal line. Here, the points are plotted 
according to the median rank method \citep{Minitabqq} to accommodate the 
right-censoring features.

```{r qqplot}
plot_qqsurv(data = firstdrink, dist = "weibull", time = "age", censor = "censor")
```

As seen in the Q-Q plot, there are some deviations from the provided $y=x$ line, 
indicating that a Weibull distribution may not be an ideal fit for the data.

## Anderson-Darling test statistic

While creating Q-Q plots are a great way to visualize how a particular distribution 
may fit the data, it can be difficult at times to definitively decide how well the 
plot fits the data. The \code{compute\_AD} function computes the Anderson-Darling (AD)
test statistic, which provides a numerical test statistic that measures how well a 
particular distribution the data fits such that lower values indicate a better fit. 
Computation of the test statistic adhered to Minitab's documentation, utilizing the 
median rank plotting method \citep{Minitabgof}.

```{r AD}
compute_AD(data = firstdrink, dist = "weibull", time = "age", censor = "censor")
```

Here, the AD test statistic value observed of 315.5693 provides no useful information 
as a stand-alone value, but rather can be used to judge fit relatively to another 
computed AD test statistic.

# Survival functions

This section provides an overview of the survival functions available in this package.
Some of the most common functions used in survival analysis are the survival function,
the hazard function, and the cumulative hazard function. Table \ref{table:functions} 
lists each function and its relationship to the other functions, as well as the 
formula used to compute each function. 


\begin{table}
\begin{tabular}{ll}
\hline
Function & Relationships  \\
\hline
PDF & ${f(t)=\frac{d}{dt}F(t)}$\\
CDF  & ${F(t)=\int_0^t f(y)dy}$\\
Survival & ${S(t)=1-F(t)=\exp[-H(t)]=\exp[-\int_0^th(y)dy]}$ \\
Hazard & ${h(t)=f(t)/S(t)=-\frac{d}{dt}\ln[S(t)]}$ \\
Cumulative hazard & ${H(t)=\int_0^t h(y)dy=-\ln[S(t)]}$\\
\hline
\end{tabular}
\label{table:functions}
\end{table}


## Survival plots

The survival function $S(t)$ estimates the proportion of subjects that survive beyond 
a specified time $t$. The \code{plot\_surv} function  plots the survival curve of 
right censored data given a specified parametric distribution.  This function also 
provides the option to plot by a grouping variable, which if specified, displays 
separate curves for each value of the specified grouping variable. In these plots, 
survival time is plotted on the $x$-axis, while survival probability is plotted on the
$y$-axis.

```{r surv}
plot_surv(data = firstdrink, dist = "weibull", time = "age", censor = "censor", by = "gender")
```

In this example, we fit a Weibull distribution to the \code{firstdrink} dataset, 
grouping by the \code{gender} variable. Here, the blue line represents the estimated 
survival curve for males (\code{group = 1}), while the red line represents the 
estimated survival curve for females (\code{group = 2}). From this plot, we see that 
the survival curve for females is consistently above the survival curve for males 
throughout all points in time. Due to this, we can conclude that males tend to 
experience their first drink of alcohol before females do.

## Hazard function

The hazard function, denoted $h(t)$, esimates the conditional risk that a subject will
experience the event of interest in the next instant of time, given that the subject
has survived beyond a certain time $t$. For example, if we observe that a person has 
survived for 17 years without first trying alcohol, the hazard function would estimate
the risk that the person will experience their first drink of alcohol in the next 
short instant of time, based on the fact that the person has already survived 17 years
alcohol free. However, hazard is a rate and not probability, and therefore can take
values greater than one.  Moreover, the hazard function can be both increasing or
decreasing.  The \code{plot\_haz} function plots the hazard function based on 
specified parametric distribution, with the option to include a grouping variable. 

```{r haz}
plot_haz(data = firstdrink, dist = "weibull", time = "age", censor = "censor",  by = "gender")
```

Based on the estimated hazard function, we see that as males and females continue to 
"survive" (not yet experience their first drink of alcohol), their risk of 
experiencing the event of interest (drinking) in the next instant of time increases.  
This risk increases more dramatically for males relative to females.


## Cumulative hazard function

While hazard plots are usually useful in assessing a subject's risk of experiencing 
the event of interest in the next moment of time, these plots can be difficult to read
and understand at times. Sometimes, the changes in hazard are very subtle, making it 
difficult to describe periods of increasing and decreasing risk. In order to 
accurately assess how hazard rates change over time, we investigate the accumulation 
of hazard rates over time, known as cumulative hazard. The cumulative hazard function,
denoted $H(t)$, is the total accumulated risk of experiencing an event up to time $t$.  

Since the cumulative hazard function is an accumulation of rates, it is important to 
note that this function is non-decreasing and is hardly ever remains constant by 
nature. The \code{plot\_cumhaz} function displays cumulative hazard plots, given that 
the data follows a specified parametric distribution. 

```{r cumhaz}
plot_cumhaz(data = firstdrink, dist = "weibull", time = "age", censor = "censor",  by = "gender")
```
As expected, the cumulative hazard function is increasing for both males and females. 
Here, the total accumulated risk of experiencing the first drink of alcohol is 
consistently greater for males compared to females. 

# Probabilities and Statistics

Survival analysis is typically accompanied by reporting summary statistics such as the
mean, median, standard deviation, and percentiles of survival time. In addition, it is
also common to compute survival probabilities.

## Survival probability

A survival probability estimates the probability that a subject survives (does not 
experience the event of interest) beyond a specified time $t$. The function 
\code{surv\_prob} computes the probability of survival beyond time $t$, given that the
data follows a specified parametric distribution. The \code{num} argument specifies
the time $t$ of interest.  The default for \code{surv\_prob} is to compute an upper
tail probability; this can be reversed to a lower tail probability useing the 
\code{lower.tail} argument.  The function can also optionally take a grouping variable
with the \code{by} argument.

```{r survprob}
surv_prob(data = firstdrink, dist = "weibull", num = 30, lower.tail = F, time = "age", censor = "censor", by = "gender")
```

Given a Weibull distribution, the overall estimated probability that a person survives
beyond 30 years without having their first drink of alcohol is approximately 0.05. 
However, males (\code{level = 1}) are less likely to survive beyond 30 years compared
to females (\code{level = 2}).

## Summary statistics

The \code{surv\_summary} function estimates various summary statistics, including
mean, median, standard deviation, and percentiles of survival time given that the data
follows a specified parametric distribution. This function also supports the option to
provide separate summary statistics for each level of a grouping variable, if desired
All summary statistics from the class \samp{fitdistcens} are provided.  If the 
distribution supplied is one of normal, lognormal, exponential, weibull, or logistic 
then the standard deviation reported is an exact computation from parameter estimates
However, if a user specifies a distribution other than that from this list, then the 
standard deviation is estimated from 1,000 randomly generated values from the 
distribution.   

```{r stats}
surv_summary(data = firstdrink, dist = "weibull", time = "age", censor = "censor", by = "gender")
```

Utilizing the grouping variable of gender, produces two separate tables of summary 
statistics for males and females, respectively. The mean survival time for males (16.2
years) is less than the mean survival time for females (18.5 years). 

# Conclusion

The \pkg{parmsurvfit} package allows for parametric survival analysis methods 
involving right-censored data to be easily computed in R. The overall goal of 
developing this package was to provide a central package for R users to implement 
typical methods used in parametric survival analysis such as computing survival 
probabilities, creating survival and hazard plots, and assessing goodness of fit of a 
parametric distribution to a dataset. 

# Acknowledgments 

This package development was funded by the Bill and Linda Frost fund at California 
Polytechnic State University, San Luis Obispo. We would also like to thank Jeff Sklar 
for his input and guidance.

\bibliography{RJreferences}